---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

  <!-- badges: start -->
  <!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

* validate objs, testa s4 klasserna  

## Installation

You can install rprimer from [GitHub](https://github.com/) with:

  ``` r
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("sofpn/rprimer")
```
``` {r, message=FALSE, warn=FALSE}
# library(rprimer)
devtools::load_all(".")
```

## Overview 

rprimer provides tools for designing broadly reactive (RT)-(q/dd)PCR assays from a multiple DNA sequence alignment. The design process is built on three functions: 

* `getConsensusProfile()`: returns an `RprimerProfile` object, which is used as input for;
* `getOligos()`: returns an `RprimerOligo` object (all the oligo candidates), which is used as input for; 
* `getAssays()`: returns an `RprimerAssay` object (all the assay candidates).

The `Rprimer`-classes are extensions of the `DataFrame` class from S4Vectors, and behave in a similar manner as traditional data frames. 

## Workflow 

### Import data 

The first step is to import an alignment with target sequences of interest and, if preferred, mask positions with e.g. high gap frequency. `readDNAMultipleAlignment()` and `maskGaps()` from Biostrings do the work for this part. 

```{r}
infile <- system.file('extdata', 'example_alignment.txt', package = 'rprimer')

myAlignment <- Biostrings::readDNAMultipleAlignment(infile, format = "fasta")
myAlignment <- Biostrings::maskGaps(myAlignment, 
                                    min.fraction = 0.5, 
                                    min.block.width = 1) 
```

### Step 1: `getConsensusProfile` 

`getConsensusProfile()` takes a `Biostrings::DNAMultipleAlignment` object as input and returns all the information needed for the subsequent design process. 

```{r}
myConsensusProfile <- getConsensusProfile(myAlignment, iupacThreshold = 0.05)
```

Output: 

```{r, echo=FALSE}
knitr::kable(head(myConsensusProfile), digits = 2)
```

The output can be visualized with `plotData()`, and specific regions can be highlighted using the optional arguments `shadeFrom` and `shadeTo`. 

```{r, fig.width=12, fig.height=6}
plotData(myConsensusProfile, shadeFrom = 5000, shadeTo = 5500)
```

### Step 2: `getOligos`

`getOligos()` searches for oligos from an `RprimerProfile`-object. All oligos are shown in both majority (without degenerate bases) and IUPAC format (with degenerate bases). 

``` {r}
myOligos <- getOligos(myConsensusProfile,
                      lengthPrimer = 18:22,
                      maxGapFrequencyPrimer = 0.05,
                      maxDegeneracyPrimer = 2,
                      gcClampPrimer = TRUE,
                      avoid3EndRunsPrimer = TRUE,
                      minEndIdentityPrimer = 1,
                      gcRangePrimer = c(0.45, 0.65),
                      tmRangePrimer = c(55, 65),
                      concPrimer = 500,
                      probe = TRUE,
                      lengthProbe = 16:24,
                      maxGapFrequencyProbe = 0.05,
                      maxDegeneracyProbe = 4,
                      avoid5EndGProbe = TRUE, 
                      gcRangeProbe = c(0.45, 0.65),
                      tmRangeProbe = c(55, 70),
                      concProbe = 250,
                      concNa = 0.05)
```

The oligo candidates can be visualized using `plotData()`:

```{r, fig.width=12, fig.height=8}
plotData(myOligos)
```

### Step 3: `getAssays`

`getAssays()` finds pairs of forward and reverse primers and, if selected, combines them with probes.

```{r}
myAssays <- getAssays(myOligos, 
                      length = 65:120,
                      maxTmDifferencePrimers = 2,
                      tmDifferencePrimersProbe = c(-2, 10))
```

The assay candidates can be visualized using `plotData()`: 

```{r, fig.width=12, fig.height=6}
plotData(myAssays)
```

## More information

The package vignette contains more information. It is loaded by `browseVignettes("rprimer")`.
